---
title: "KEGG Orthology Network Analyzer"
author: "fdelogu"
date: '`r format(Sys.time(), "%d %B, %Y")`'
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "results") })
output:
  html_document:
    self_contained: no
    toc: true
    toc_float: true
    theme: united
---

# KEGG Orthology Network Analyzer

## Getting started

### Set options and load libraries

```{r load libraries and settings, message=F}

options(stringsAsFactors=F)
wd <- paste0(getwd(), "/")

library(tidyverse)
library(igraph)

```

### Load premade maps and the adjacency matrix

```{r load data}

if(file.exists(paste0(wd, "results/ko.txt"))){
  KOmap.df <- read.table(paste0(wd, "results/ko.txt"), header=T, sep="\t")
} else {
  print("You should run ko_hierarchy_parser.Rmd first.")
}

if(file.exists(paste0(wd, "results/ko2cko.txt"))){
  ko2cko <- read.table(paste0(wd, "results/ko2cko.txt"), header=T)
} else {
  print("You should run net_maker.Rmd first.")
}

if(file.exists(paste0(wd, "results/CKO_Adj.txt"))){
  adj <- data.matrix(read.table(paste0(wd, "results/CKO_Adj.txt")))
} else {
  print("You should run net_maker.Rmd first.")
}

```

## Topological analysis

### Degree distribution

Compute the degree distribution of the metabolic network:

```{r compute degree}

deg <- colSums(adj)
top10nodes <- sort(deg, decreasing=T)[1:10]

top10nodes

```

Map the CKOs on the original KOs

```{r find corresponding KOs}

explosion <- function(df, col1, x){
  return(df[df[,col1]==x,])
}

ko2cko.subset <- data.frame(t(rep("", ncol(ko2cko))))
colnames(ko2cko.subset) <- colnames(ko2cko)
for(i in names(top10nodes)){
  ko2cko.subset <- rbind(ko2cko.subset, explosion(ko2cko, "CKO", i))
}
ko2cko.subset <- ko2cko.subset[-1,]

ko2cko.subset

```

Map the retrieved KOs on the names form the Hierarchy:

```{r find KOs names}

concussion <- function(df1, df2, col1){
  
  df3 <- 1:(ncol(df1)+ncol(df2)-1)
  
  inter_index <- unique(df1[,col1])[unique(df1[,col1])%in%unique(df2[,col1])]
  
  for(i in inter_index){
    
    left_block <- as.data.frame(explosion(df1, col1, i))
    right_block <- as.data.frame(explosion(df2, col1, i)[,colnames(df2)[colnames(df2)!=col1]])
    
    for(j in 1:nrow(left_block)){
      for(z in 1:nrow(right_block)){
        df3 <- rbind(df3, t(c(unlist(left_block[j,]), unlist(right_block[z,]))))
      }
    }
    
  }
  
  colnames(df3) <- c(colnames(df1), colnames(df2)[colnames(df2)!=col1])
  
  
  return(as.data.frame(df3[-1,]))
}

to_map <- ko2cko.subset %>%
  separate(KO, c("ko", "D_code"), 3) %>%
  select(-ko)

mapped_df <- concussion(KOmap.df, to_map, "D_code")

mapped_df %>%
  group_by(D_name, CKO) %>%
  summarise(n=n())

rm(ko2cko.subset, to_map, mapped_df)
```

### Average degree of different metabolisms

We want to use the netwrok to compute the average degree for each metabolism compared to the number of nodes they have. Subset the *KOmap* to the metabolic only entries, then map it on the CKOs and take the intersection.

```{r Subset KOmap to Metabolism only}

KOmap.metabolism <- KOmap.df %>%
  filter(A_name=="Metabolism")

deg.df <- data.frame(DEG=deg, CKO=names(deg))

ko2cko.subset <- concussion(ko2cko, deg.df, "CKO")

to_map <- ko2cko.subset %>%
  separate(KO, c("ko", "D_code"), sep=":") %>%
  select(-ko)

deg2KOmap <- concussion(to_map, KOmap.metabolism, "D_code")

dim(deg2KOmap)

```

Plot the everage degree vs. the number of nodes for each metabolism.

```{r Plot average degree vs #nodes per metabolism pathway}

plt1 <- deg2KOmap %>%
  group_by(B_name, C_name) %>%
  summarise(n=n(), avr_deg=mean(as.numeric(DEG))) %>%
  ggplot(aes(x=n, y=avr_deg, color=B_name)) +
  geom_point() +
  labs(x="Number of nodes", y="Average degree", title="Whole KEGG network") +
  guides(color=guide_legend(title="Metabolism")) +
  geom_smooth(method="lm", se=F) +
  theme_classic()

rm(deg2KOmap, to_map, ko2cko.subset, KOmap.metabolism)
ggsave(paste0(wd, "results/kegg_NDeg.png"), plt1, width=10, height=6, dpi=300)
plt1

```

## Plot the network

### Simple plot

Plot the network using the degree of the nodes to set their size.

```{r Plot network}

G <- graph_from_adjacency_matrix(adj, mode=c("undirected"), weighted = NULL, diag = TRUE)

table(table(components(G)$membership)) # Number of connected componenets per number of nodes in it

to_map <- ko2cko %>%
  separate(KO, c("ko", "D_code"), sep=":") %>%
  select(-ko)

CKO2KOmap <- concussion(to_map, KOmap.df, "D_code")

G1.names <- names(components(G)$membership)[components(G)$membership==1] # Isolate the largest component
G1 <- graph_from_adjacency_matrix(adj[G1.names,G1.names], mode=c("undirected"), weighted = NULL, diag = TRUE)
plot(degree.distribution(G1))

l <- layout_with_kk(G1)
V(G1)$size <- log10(degree(G1))*5
V(G1)$shape <- "circle"

plot(G1, edge.arrow.size=.0, vertex.label=NA, layout=l)

```

### Plot and annotation

Color the nodes using the information in the second level of the KEGG Pathways and Brite Hierarchy (*B_name*). Nodes with more than one label will have more than one color. If the node does not have a term will be grey.

```{r Plot network and annotation}

A_class <- as.factor(c(unique(CKO2KOmap$B_name), "Unclassified"))  # HERE
tmp <- CKO2KOmap %>%
  group_by(CKO, B_name) %>%  # HERE
  summarise(n=n())
color_list <- as.list(rep("", length(unique(ko2cko$CKO))))
names(color_list) <- unique(ko2cko$CKO)
tmp_holder <- unique(CKO2KOmap$CKO)
for(i in unique(ko2cko$CKO)){
  if(i %in% tmp_holder){
    a <- rep(0, length(A_class))
    names(a) <- levels(A_class)
    b <- explosion(tmp, "CKO", i)
    a[factor(b$B_name, levels=levels(A_class))] <- b$n   # HERE
  } else {
    a <- c(rep(0, (length(A_class)-1)), 1)
    names(a) <- levels(A_class)
  }
  color_list[[i]] <- as.vector(a) #/sum(as.vector(a))
}

nodes_shares <- unlist(lapply(color_list, function(x) sum(ifelse(x>0, 1, 0))==1))
pure_nodes <- names(nodes_shares)[nodes_shares]

png(paste0(wd, "results/net_plot.png"))
plot(G1, edge.arrow.size=.0, vertex.label=NA, layout=l,
     vertex.shape="pie",  vertex.pie=color_list[names(V(G1))],
     vertex.pie.color=list(c(heat.colors(length(A_class)-1), "grey")),
     )
dev.off()

plot(G1, edge.arrow.size=.0, vertex.label=NA, layout=l,
     vertex.shape="pie",  vertex.pie=color_list[names(V(G1))],
     vertex.pie.color=list(c(heat.colors(length(A_class)-1), "grey")),
     )

# vertex.shape=c("pie", "circle")[(names(V(G1))%in%pure_nodes)+1], 

```
